---
title: "ANLY 565 Time Series and Forecasting (Project Carbon Emission)"
author: "Sonachi Mogbogu, Uttara Suryavanshi, Rahul Awale"
date: "`r Sys.Date()`"
output:

  html_document:
    df_print: paged
---



```{r}

   library(readr)
   library(ggplot2)
   library(forecast)
   library(fpp2)
   library(TTR)
   library(dplyr)
   library(mice)
   library(readxl)
   library(moments)
   library(ggpubr)
   library(tseries)

  # if (!exists(.Random.seed)) runif(1)
  #    seed <- .Random.seed

   set.seed(1000)      



```

1  Check you working directory
```{r}
   getwd()

```

2  Set your working directory to file path 
```{r}
   setwd('C:/Users/obihov/Documents/ANLY 565/RScript/project')
   print(getwd())
  
```

3  Open file and fix the date column format to follow "YYYY-MM-DD"
```{r}

   carbon_emission <- read.csv("MER_T12_06.csv", header = TRUE)
 
   carbon_emission$Value <- as.numeric(as.character(carbon_emission$Value))
   carbon_emission$YYYYMM <- paste0(carbon_emission$YYYYMM, "01")
   carbon_emission$YYYYMM  <- as.Date(as.character(carbon_emission$YYYYMM), '%Y%m%d')   
```

4. Spot missing values and conduct reprocessing of data


```{r viz}
   library(visdat)
   
   vis_dat(carbon_emission)
   
   vis_miss(carbon_emission)
   
   summary(carbon_emission)

```

```{r}

   # viewing table
   glimpse(carbon_emission)

   # accuracy check
   summary(carbon_emission)
   
   # Checking and resolving missing data
   apply(carbon_emission,2,function(x) sum(is.na(x)))
   
   ## resolving with mean substitution
   percentmiss = function(x){sum(is.na(x))/length(x)*100}
   apply(carbon_emission,1,percentmiss)
   missing = apply(carbon_emission,1,percentmiss)
   table(missing)
   replace = subset(carbon_emission, missing <= 5)
   missing0 = apply(replace,1,percentmiss)
   table(missing0)
   
   ## subsetting
   dont = subset(carbon_emission, missing > 5)
   missing2 = apply(dont,1,percentmiss)
   table(missing2)
   
   apply(carbon_emission,2,percentmiss)
   
   # excluding categorical variables
   replace_col = replace[,-c(1,3)]
   dont_col = replace[,c(1,3)]
   

   temp_no_miss = mice::mice(replace_col) 
   
   no_miss = complete(temp_no_miss,1)
   summary(no_miss)
   
   no_na_carbon_emission = cbind(dont_col, no_miss)
   summary(no_na_carbon_emission)
   
   vis_dat(no_na_carbon_emission)

   vis_miss(no_na_carbon_emission)
   
```


```{r}

   hist(no_na_carbon_emission$Value, xlab = 'carbon emission value', breaks = 15)

   # Before Transformation
   skewness(no_na_carbon_emission$Value, na.rm=TRUE)
   kurtosis(no_na_carbon_emission$Value, na.rm=TRUE)

   #testing normality
   shapiro.test(no_na_carbon_emission$Value)
   
   # Transformation
   
   no_na_carbon_emission$Value <- log10(no_na_carbon_emission$Value+1 - min(no_na_carbon_emission$Value))
   
   # After Transformation
   
   skewness(no_na_carbon_emission$Value, na.rm=TRUE)
   kurtosis(no_na_carbon_emission$Value, na.rm=TRUE)
    
   ggdensity(no_na_carbon_emission, x = "Value", 
             fill = "lightgray", title = "CONT") +
   stat_overlay_normal_density(color = "red", linetype = "dashed")
   
   hist(no_na_carbon_emission$Value, xlab = 'carbon emission value', breaks = 15) # multinomial distribution

   #normality
   shapiro.test(no_na_carbon_emission$Value)

```


```{r}

   # Recoding categorical data

   no_na_carbon_emission <- no_na_carbon_emission %>% 
   mutate(
       description_reduced = recode(
          .$Description,
          "Coal Electric Power Sector CO2 Emissions" = "coal",
          "Distillate Fuel, Including Kerosene-Type Jet Fuel, Oil Electric Power Sector CO2 Emissions" = "distillate fuel",   
          "Geothermal Energy Electric Power Sector CO2 Emissions"="geothermal energy",
          "Natural Gas Electric Power Sector CO2 Emissions" = "natural gas",      
          "Non-Biomass Waste Electric Power Sector CO2 Emissions"="non-biomass waste",      
          "Petroleum Coke Electric Power Sector CO2 Emissions" = "petroleum coke",      
          "Petroleum Electric Power Sector CO2 Emissions" = "petroleum electric power",      
          "Residual Fuel Oil Electric Power Sector CO2 Emissions" = "residual fuel oil", 
          "Total Energy Electric Power Sector CO2 Emissions" = "total energy"
       )
    )

   
   # tabulate results
   
   table(no_na_carbon_emission$description_reduced)

```


```{r, warning=FALSE, message=FALSE}

   boxplot<-ggplot(data=no_na_carbon_emission) +
   geom_boxplot(mapping=aes(x=description_reduced, y=Value, color=description_reduced)) +
   labs(title="Carbon Emission in the United States", y="Carbon Emission(in million metric tons)", x="")
   
boxplot<-boxplot+guides(color=guide_legend(title="Type"))
   
   boxplot


```

```{r, warning=FALSE, message=FALSE}
   #histogram
   
   hist1<-ggplot(data=no_na_carbon_emission, aes(Value))+labs(title = "Carbon Emission Data Distribution by Value in the United States",x="Carbon Emission(in million metric tons)", y="Frequency")+geom_histogram(color="white",fill="gold", bins=10)
   
   hist1
   
   hist2<-ggplot(data=no_na_carbon_emission, aes(YYYYMM))+labs(title = "Carbon Emission Data Distribution by Date in the United States",x="Date", y="Frequency")+geom_histogram(color="white",fill="sky blue", bins=43)
   
   hist2
   
   hist3<-ggplot(data=no_na_carbon_emission, aes(x=Column_Order), stat_count(Column_Order))+labs(title = "Carbon Emission Data Distribution by Date in the United States",x="Type of Emission", y="Frequency")+geom_histogram(color="white",fill="gray", bins=9)
   
   hist3
   
   
```
   

```{r corr, warning=FALSE, message=FALSE}

   #creating separate values 
   
   date <- no_na_carbon_emission$YYYYMM
   
   emission<-as.numeric(no_na_carbon_emission$Value)
   
   type_no<-as.numeric(no_na_carbon_emission$Column_Order)
   
   type_name<-no_na_carbon_emission$description_reduced
   
   emission_date<-cbind.data.frame(emission,date)
   
   emission_type<-cbind.data.frame(emission,type_no)
   
   emission_date_type<-cbind.data.frame(emission,date,type_no)
   
   emission_date_type_name<-cbind.data.frame(emission,date,type_no,type_name)
   #checking correlation
   
   cor(emission_type, method = c("pearson","kendall","spearman"))
   
   #creating a function
   flattenCorrMatrix <- function(cormat, pmat) {
     ut <- upper.tri(cormat)
     data.frame(
       row = rownames(cormat)[row(cormat)[ut]],
       column = rownames(cormat)[col(cormat)[ut]],
       cor  =(cormat)[ut],
       p = pmat[ut]
       )
   }
   
   library(Hmisc)
   library(corrplot)
   
   res<-rcorr(as.matrix((emission_type)))
   res
   
   flattenCorrMatrix(res$r, res$P)
   
   # Insignificant correlations are leaved blank
   corrplot(res$r, type="upper", order="hclust", 
            p.mat = res$P, sig.level = 0.01, insig = "blank")
   
   library("PerformanceAnalytics")
   chart.Correlation(emission_type, histogram = TRUE, pch=19)
   
   #no correlation between type and carbon emission value


```

```{r bar, warning=FALSE, message=FALSE}

   barchart1<-ggplot(data=emission_date_type_name, aes(x=emission, y=type_name, fill=type_name))+
      labs(title="Carbon Emission in the United States", x="Carbon Emission(in million metric tons)", y="")+geom_col(position = "dodge")
   barchart1<-barchart1+guides(fill=guide_legend(title="Type"))
   
   barchart1
   
   barchart2<-ggplot(data=emission_date_type_name, aes(x=date, y=emission, fill=type_name))+
      labs(title="Carbon Emission in the United States", x="Carbon Emission(in million metric tons)", y="")+geom_col(position = "dodge")
   barchart2<-barchart2+guides(fill=guide_legend(title="Type"))
   
   barchart2
  
```

```{r scatter, warning=FALSE, message=FALSE}
   scatterplot1<-ggplot(data=emission_date_type_name, aes(x=date, y=emission))+
      labs(title="Carbon Emission in the United States", x="Date", y="Carbon Emission(in million metric tons)")+
      geom_point(aes(color=type_name))+theme(axis.text.x=element_text(angle=90,hjust = 1,vjust = 1))
   
   scatterplot1

```

```{r line, warning=FALSE, message=FALSE}
   linechart1<-ggplot(emission_date_type_name, aes(x=date, y=emission))+geom_line(aes(color=type_name))+
      labs(title="Carbon Emission in the United States", x="Date", y="Carbon Emission(in million metric tons")+theme(axis.text.x=element_text(angle=90,hjust = 1,vjust = 1))
   
   linechart1<-linechart1+guides(color=guide_legend(title="Type"))
   
   linechart1
                                 
```

# Time series Analysis with Arima or Exponential Smoothing

```{r}

   carbon_emission_value <- no_na_carbon_emission
   range(date)

   apply(carbon_emission_value, 2, function(x) any(is.na(x))) # checking for missing values

```

```{r timeseries2}
  
  ts_coal <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='coal')], start = c(1973, 1), end=c(2016,7), frequency = 12 )
  ts_distillateFuel <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='distillate fuel')], start = c(1973, 1),  end=c(2016,7), frequency = 12 )
  ts_geothermalEner <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='geothermal energy')], start = c(1973, 1),  end=c(2016,7),frequency = 12 )
  ts_naturalGas <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='natural gas')], start = c(1973, 1),  end=c(2016,7),frequency = 12 )
  ts_nonBiomass <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='non-biomass waste')], start = c(1973, 1), end=c(2016,7), frequency = 12 )
  ts_petroleumCoke <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='petroleum coke')], start = c(1973, 1),  end=c(2016,7),frequency = 12 )
  ts_petroleumElectric <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='petroleum electric power')], start = c(1973, 1),  end=c(2016,7), frequency = 12 )
  ts_residualFuel <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='residual fuel oil')], start = c(1973, 1),  end=c(2016,7),frequency = 12 )
  ts_totalEner <- ts(carbon_emission_value$Value[which(no_na_carbon_emission$description_reduced=='total energy')], start = c(1973, 1),  end=c(2016,7),frequency = 12 )
  ts_allEmission <- ts(carbon_emission_value$Value, start = c(1973, 1), end=c(2016,7), frequency = 12)
  
  plot(ts_coal, main = "Carbon Emission for Coal from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value")
  plot(ts_distillateFuel, main = "Carbon Emission for distillate fuel from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value" )
  plot(ts_geothermalEner, main = "Carbon Emission for geothermal energy from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value")
  plot(ts_naturalGas, main = "Carbon Emission for geothermal energy from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value")
  plot(ts_nonBiomass, main = "Carbon Emission for geothermal energy from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value")
  plot(ts_petroleumCoke, main = "Carbon Emission for petroleum coke from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value")
  plot(ts_petroleumElectric, main = "Carbon Emission for petroleum electric power from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value")
  plot(ts_residualFuel, main = "Carbon Emission for residual fuel oil from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value")
  plot(ts_totalEner, main = "Carbon Emission for total energy from Jan 01, 1973 - July 01, 2016", xlab = "Year", ylab = "Carbon emission value")

```
```{r adf.test}

  adf.test(ts_allEmission)

```
__You can see the box plots leaving the 95 percent confidence boundary crossing the threshold from the autocorrelation. From this, it is to note that there is autocorrelation and resulting in data being non-stationary. To make the data to be stationary, enforcing the difference concept is the best approach, and from there, we can carry out more operations. We would be using total_energy. P-value is  0.9828, which is more than the significance level of 0.05. Therefore we fail to reject the null hypothesis.__ 

# Detrending technique is used to remove the trend component from the time series to make it stationary 

```{r}

  ts_allEmission_stl <- stl(ts_allEmission, s.window = "period")$time.series[,2]
  
  detrend_ts = ts_allEmission - (ts_allEmission_stl - ts_allEmission_stl[1])
  plot(detrend_ts)
  
  ts_allEmission <- detrend_ts
  
  adf.test(ts_allEmission)
  

  acf(ts_allEmission) #autocorellation
  pacf(ts_allEmission)  #partial autocorellation

```

__Here, the p-value is 0.01, which is less than the significance level, so we have enough evidence to reject the null hypothesis. It appears to be stationary with constant mean and variance and having a unit root. Although there seems to be autocorrelation present with the ACF showing eight box plots going outside of the 95 percent confidence boundary.__ 

```{r }

   carbonEmission_train <- window(ts_allEmission, start = c(1990,1), end = c(2015, 7), 
                     frequency = 12)
   carbonEmission_test <- window(ts_allEmission, start = c(2015,8), end = c(2016,7), 
                      frequency = 12)
   
```

```{r 1, warning=FALSE, message=FALSE}

   Time<-time(carbonEmission_train)
   
   Seas<- factor(cycle(carbonEmission_train))
   
   
   carbonEmission.lm<- lm(carbonEmission_train~0+Time+Seas)
   
   summary(carbonEmission.lm)

```

```{r 2, warning=FALSE, message=FALSE}

   new.Time<- seq(2015.55, len= 12, by= 1/12)
   
   new.Seas<- factor(c(7,8,9,10,11,12,1,2,3,4,5,6))
   
   new.data<- data.frame(Time=new.Time, Seas=new.Seas)
   
   new.data

```

```{r predict, warning=FALSE, message=FALSE}

   predict.lm<-predict(carbonEmission.lm, new.data)
   
   predict.lm

```

```{r resid, warning=FALSE, message=FALSE}

   carbonEmission.lm.resid<- carbonEmission.lm$residuals
   
   acf(carbonEmission.lm.resid)
   
   pacf(carbonEmission.lm.resid)

```

```{r best order, warning=FALSE, message=FALSE}

   best.order <- c(0, 0, 0)
   
   best.aic <- Inf
   for (i in 0:3) for (j in 0:3) {
     fit.aic <- AIC(arima(carbonEmission.lm.resid, order = c(i, 0,
                                                  j)))
     if (fit.aic < best.aic) {
       best.order <- c(i, 0, j)
       best.arma <- arima(carbonEmission.lm.resid, order = best.order)
       best.aic <- fit.aic
     }}
   
   best.order

```

```{r coff, warning=FALSE, message=FALSE}

   resid.best.arma <- best.arma
   
   coef(best.arma)

```

```{r, warning=FALSE, message=FALSE}

   resid.best.arma.pred <- predict(resid.best.arma, n.ahead = 12)
   
   resid.best.arma.pred$pred

```

```{r, warning=FALSE, message=FALSE}

   carbonEmission.pred<- ts((predict.lm+resid.best.arma.pred$pred),start=c(2015,7), frequency = 12)
   
   ts.plot(cbind(carbonEmission_train, carbonEmission.pred), lty= 1:2)
   
   ts.plot(cbind(carbonEmission.pred, carbonEmission_test), lty=1:2)

```

```{r error, warning=FALSE, message=FALSE}

   #Means absolute error for forecasting
   
   mean(abs(carbonEmission_test - carbonEmission.pred)/carbonEmission_test)*100
   
   ts_allEmission<-na.omit(ts_allEmission)
   
   summary(ts_allEmission)
   
```

```{r}

   #using naive bayes for forecasting
   
   library(forecast)
   library(fpp2)
   library(TTR)
   library(dplyr)

```

```{r}

   head(carbonEmission_train)
   
   head(carbonEmission_test)

```

```{r}

   mape <- function(actual,pred){
    mape <- mean(abs((actual - pred)/actual))*100
    return (mape)
   }

```

```{r}

   naive_mod <- naive(carbonEmission_train, h = 12)
   summary(naive_mod)
   naive_df<-as.data.frame(naive_mod)
   
   carbonemission_testdf<-as.data.frame(carbonEmission_test)
   carbonemission_testdf$naive<-naive_df$`Point Forecast`
   
   naive_mape=mape(carbonemission_testdf$x,carbonemission_testdf$naive)
   naive_mape
   
```

```{r}

   #Simple Exponential Smoothing
   
   se_model <- ses(carbonEmission_train, h = 12)
   summary(se_model)
   
   se_df<-as.data.frame(se_model)
   
   carbonemission_testdf$se<-se_df$`Point Forecast`
   
   se_mape=mape(carbonemission_testdf$x,carbonemission_testdf$se)
   se_mape
   
```

```{r}

   #holt's trend method
   
   holt_model <- holt(carbonEmission_train, h = 12)
   
   summary(holt_model)
   
   df_holt = as.data.frame(holt_model)
   
   carbonemission_testdf$holt<-df_holt$`Point Forecast`
   
   holt_mape=mape(carbonemission_testdf$x,carbonemission_testdf$holt)
   holt_mape
   
```

```{r}

   arima_model <- auto.arima(carbonEmission_train)
   
   summary(arima_model)
   
   fore_arima = forecast::forecast(arima_model, h=12)
   
   df_arima = as.data.frame(fore_arima)
   
   
   carbonemission_testdf$arima<-df_arima$`Point Forecast`
   
   arima_mape= mape(carbonemission_testdf$x,carbonemission_testdf$arima)
   arima_mape

```

```{r}

   model_tbats <- tbats(carbonEmission_train)
   summary(model_tbats)
   
   fore_tbats = forecast::forecast(model_tbats, h=12)
   
   df_tbats <-as.data.frame(fore_tbats)
   
   carbonemission_testdf$tbats<-df_tbats$`Point Forecast`
   
   tbats_mape = mape(carbonemission_testdf$x,carbonemission_testdf$tbats)
   
   tbats_mape
   
```

```{r mape}

   mape_Score_df = data.frame("Model_name" = c('Naive', 'Exponential Smoothing', 'Holts Trend', 'Arima', 'TBATS'),
                           "Accuracy_Score" = c(naive_mape, se_mape, holt_mape, arima_mape, tbats_mape)) 
     
   # display dataframe
   head(mape_Score_df)

```

```{r mapeVisualization}

   # Add title, narrower bars, fill color, and change axis labels
   ggplot(data=mape_Score_df, aes(x=Model_name, y=Accuracy_Score, fill=Model_name)) + 
       geom_bar(width=.8, stat="identity") + 
       guides(fill=FALSE) +
       xlab("Model name") + ylab("Accuracy score: MAPE") +
       ggtitle("Accuracy metrics result") +
       geom_bar(stat="identity") +
          scale_fill_manual(values=c("red2", "slateblue4", "green3", "orange", "olivedrab2")) + theme_bw()
    
    
```

__The Naive, Simple Exponential Smoothing, and Holts Trend models did well at equally achieving a lower MAPE of 6.28 percent, 4.12 percent, and 4.16 percent, respectively. However, ARIMA and TBATS model emerged as the winner based on their performances on the test data with MAPE close to 2.31 percent and 2.60 percent, respectively. Therefore, we would use the Arima model to predict the next four years. __


```{r four years}

   myforecast<-forecast(arima_model,level=c(95),h=5*12)
   
   myforecast
  
   plot(myforecast) 
   
   #carbon emission of july 2016 vs july 2020

   percent_change= ((1.906860-1.803964)/1.803964)*100
   
   percent_change

  
```

